A Generic and Efficient E-field Parallel Imaging Correlator for Next-Generation Radio Telescopes

### 2017-01-10 imaging_demo/ms.tex
Remove strikethroughs and bold fonts
From c6789d1 (mnras):
====================

In ms.tex,

(*) Remove strikethroughs and bold fonts

% Only include extra packages if you really need them. Common packages are:
\usepackage{graphicx}	% Including figure files
\usepackage{amsmath}	% Advanced maths commands
\usepackage{amssymb}	% Extra maths symbols
\usepackage[flushleft]{threeparttable} % For table notes
\usepackage[caption=false]{subfig} % For subfloats
\usepackage[normalem]{ulem} % For strikethrough text

@@

% Abstract of the paper
\begin{abstract}
Modern radio telescopes are favouring densely packed array layouts with large numbers of antennas ($\Nant\gtrsim 1000$). Since the complexity of traditional correlators scales as $\mathcal{O}(\Nant^2)$, there will be a steep cost for realizing the full imaging potential of these powerful instruments. Through our generic and efficient E-field Parallel Imaging Correlator (EPIC), we present the first software demonstration of a generalized direct imaging algorithm, namely, the Modular Optimal Frequency Fourier (MOFF) imager. \sout{It takes advantage of the multiplication-convolution theorem of Fourier transforms.} Not only does it bring down the cost for dense layouts to $\mathcal{O}(\Nant\log_2\Nant)$ but can also image from irregular layouts and heterogeneous arrays of antennas. EPIC is highly modular, parallelizable, implemented in object-oriented Python, and publicly available. We have verified the images produced to be equivalent to those from traditional techniques to within a precision set by gridding coarseness. We have also validated our implementation on data observed with the Long Wavelength Array (LWA1). {\bf We provide a detailed framework for imaging with heterogeneous arrays and show that EPIC} robustly estimates the input sky model for {\bf such} arrays. Antenna layouts with dense filling factors consisting of a large number of antennas such as LWA, the Square Kilometre Array, Hydrogen Epoch of Reionization Array, and Canadian Hydrogen Intensity Mapping Experiment will gain significant computational advantage by deploying an optimized version of EPIC. The algorithm is a strong candidate for instruments targeting transient searches of Fast Radio Bursts (FRB) as well as planetary and exoplanetary phenomena due to the availability of high-speed calibrated time-domain images and low output bandwidth relative to visibility-based systems.
\end{abstract}

@@

Here we present the first software implementation of the MOFF correlator, and announce the public release of the E-field Parallel Imaging Correlator (EPIC) code. EPIC is a highly parallel, object-oriented Python package that primarily implements the MOFF imaging algorithm, \sout{as well as emulating} {\bf emulates} real-life telescopes and FX/XF correlators in software, and includes a visibility-based imaging technique for reference. It is intended to provide a development platform to test different imaging approaches, characterize scaling relations and serve as a stepping stone for real-life GPU/FPGA-based implementation on telescopes.

We begin with a technical description of the algorithm in \S\ref{sec:math}, then discuss our particular implementation in \S\ref{sec:software}. We then verify the output data quality from our code in \S\ref{sec:verify} by presenting simulated images from both the EPIC correlator and comparing to a simulated FX correlator. We also demonstrate the performance with real-world data from the LWA1. {\bf In \S\ref{sec:versatility}, we present a mathematical framework for imaging with heterogeneous arrays and demonstrate the ability of EPIC} to robustly image data from such arrays. In \S\ref{sec:analysis}, we explore the scalability of the algorithm in the context of several array design choices. We identify specific array design classes where the EPIC correlator is computationally more efficient, and in the field of transients, demands significantly lesser I/O bandwidth relative to visibility-based approaches. We conclude and discuss future research prospects in \S\ref{sec:conclusions}.

@@

We can then use the multiplication-convolution theorem to move the convolution in Equation~\ref{eqn:e-field-conv} to a \sout{square} {\bf product} after the Fourier transform in equation~\ref{eqn:dirty-image-FX}.
\begin{align}
  \boldsymbol{\mathcal{I}}^\prime(\hat{\mathbf{s}}) &= \left\langle \left|\,\mathbf{F}^\textrm{T}(\hat{\mathbf{s}},\mathbf{r})\,\mathbf{W}^\textrm{T}(\mathbf{r},\mathbf{a})\,\widetilde{\mathbf{N}}(\mathbf{a},\mathbf{a})\,\mathbf{E}(\mathbf{a})\,\right|^2\right\rangle_t. \label{eqn:dirty-image-MOFF}
\end{align}
The term inside the angular brackets before squaring has a very similar form as that in equation~\ref{eqn:dirty-image-FX}. It signifies that the measured antenna electric fields are weighted by the antenna noise, weighted and gridded by the antenna aperture kernel, Fourier-transformed and finally squared to obtain the same image estimate that would have been obtained using equation~\ref{eqn:dirty-image-FX}.

@@

In a companion paper \citep{bea16}, we demonstrate a novel calibration technique (EPICal) which leverages the data products formed by direct imaging correlators to estimate antenna complex gains. This method correlates the antenna electric field signals with an image pixel from the output of the correlator in the feedback calibration fashion outlined in \citealt{mor11} (illustrated in Fig.~\ref{fig:MOFF-flowchart} by the arrow leading from the imager to the calibration block). Furthermore, it allows for arbitrarily complex sky models, \sout{and following the MOFF algorithm} places no restrictions on array layout, and accounts for non-identical antenna beam patterns. Direction-dependent calibration can be achieved by correlating antenna signals with output pixels in the direction of $N_\textrm{c}$ calibration sources, then fitting for a functional model of the sky. Since antennas are only correlated with calibrator pixels, the computational complexity scales as $\sim N_\textrm{a} N_\textrm{c}$. The calibration performance of EPIC is explored for varying levels of completeness of sky model in \citet{bea16}.

@@

In contrast, visibilities measured by an antenna are statistically stable within an integration time interval. Hence, they are averaged after {\bf (and sometimes before)} calibration as shown in equation~\ref{eqn:cc-vis}. It is advantageous to average them in visibilities before imaging because the repeated cost of spatial FFT can be avoided. Since this averaging has been performed already on the visibilities over an integration time-scale, the imaging step has to be performed only once per integration cycle.

@@

We use the EPIC simulator to generate stochastic electric field samples from a sky model consisting of 10 point sources of random flux densities $>10$~Jy each at random locations. In our simulations, we use 64 frequency channels each of width $\Delta f = 40$~kHz. The number of time stamps integrated in one integration cycle was kept at eight where {\bf the duration of} each \sout{A/D} timeseries is $1/\Delta f=25\,\mu$s long. We use the MWA array layout \citep{bea12} for demonstration. Only the inner 51 tiles within a square bounding box of 150~m on each side were used. We assumed all tiles are identical and have a square-shaped electric field illumination footprint 4.4~m on each side. Besides the stochastic sky noise present in the simulated electric fields, no noise from the instrument is added.

@@

For a heterogeneous array, 
\begin{align}\label{eqn:wt-dirty-image-EPIC-decomp}
  \mathcal{I}^\prime &= \frac{1}{\Nant(\Nant-1)}\,\sum_{pq}\sum_{\substack{ab\\a\ne b\\a\in p, b\in q}}\,\mathcal{I}^\prime_{ab},
\end{align}
where, $ab$ indexes pairs of antennas $a$ and $b$, {\bf $pq$ indexes the different types of antenna pairs, with $p$ and $q$ indexing antenna types (note that $ab$ and $pq$ are simply indices in our notation and are not to be interpreted as products of the individual indices)}. Thus, equation~\ref{eqn:wt-dirty-image-homogeneous} is re-written as the sum of fringes over all unique antenna pairs in an antenna pair type (inner sum) and subsequently summed over all antenna pair types (outer sum).

For purposes of demonstration in this study, we assume that the spatial distribution of antenna pairs in each antenna pair type is similar and thus results in a similar dirty image, $\mathcal{I}_\textrm{D}^\textrm{iso}$. This usually holds when antennas under each antenna type are chosen randomly or identically such that differences in the synthesized beams are insignificant. Then, equation~\ref{eqn:wt-dirty-image-EPIC-decomp} can be approximated as:
\begin{align}\label{eqn:wt-dirty-image-EPIC-decomp-approx1}
  \mathcal{I}^\prime &\approx \frac{\mathcal{I}_\textrm{D}^\textrm{iso}}{\Nant(\Nant-1)}\,\sum_{pq}\sum_{\substack{ab\\a\ne b\\a\in p, b\in q}} \mathcal{W}^{\textrm{G}\star}_a\,\mathcal{W}^\textrm{G}_b\,\mathcal{W}^\textrm{I}_a\,\mathcal{W}^{\textrm{I}\star}_b,
\end{align}
where, $\mathcal{I}_\textrm{D}^\textrm{iso}$ is the same for all terms and has been pulled out of the summations. For all antennas indexed by $a$ that belong to a particular type $p$, we replace $\mathcal{W}_a$ with $\mathcal{W}_{(p)}$ and is applicable to those arising from both instrumental (superscript `$I$') and gridding (superscript `$G$') origins. If $n_{pq}$ is the number of antenna pairs in the antenna pair type $pq$ {\bf (unrelated to the noise quantity also denoted by $n$ in \S\ref{sec:math})}, $\mathcal{W}_\textrm{eff}$ can be simplified as:
\begin{align}\label{eqn:wt-dirty-image-EPIC-decomp-approx2}
  \mathcal{W}_\textrm{eff} &\approx \frac{1}{\Nant(\Nant-1)}\,\sum_{pq}n_{pq}\,\mathcal{W}^{\textrm{G}\star}_{(p)}\,\mathcal{W}^\textrm{G}_{(q)}\,\mathcal{W}^\textrm{I}_{(p)}\,\mathcal{W}^{\textrm{I}\star}_{(q)}.
\end{align}
In an optimally weighted image, $\mathcal{W}^\textrm{G}_{(p)} \equiv \mathcal{W}^\textrm{I}_{(q)}$. Thus,
\begin{align}\label{eqn:effective-weighting-optimal}
  \mathcal{W}_\textrm{eff}^\textrm{opt} &\approx \frac{1}{\Nant(\Nant-1)}\,\sum_{pq} n_{pq}\,\left|\mathcal{W}^\textrm{I}_{(p)}\right|^2\left|\mathcal{W}^\textrm{I}_{(q)}\right|^2,
\end{align}
where, superscript `$\textrm{opt}$' denotes `optimal'.

@@

The number of complex multiplications and \sout{additions} {\bf accumulations} in the spatial Fourier transform implemented via FFT is $\approx \beta\,(4\Ngrid)\log_2(4\Ngrid)$ where $\Ngrid$ is the number of pixels on the grid, the factor 4 accounts for the increase in the number of pixels as a result of zero-padding before the spatial Fourier transform, and $\beta$ is a constant that depends on the implementation of twiddle FFT algorithms \citep{bri74}. In our study, we set $\beta=5$, a value\footnote{http://www.fftw.org/speed/method.html} much more conservative than was indicated in \citet{mor11}. We set the number of complex multiplications in the X-engine in visibility-based imaging to $\Nant(\Nant-1)/2$.

@@

Fig.~\ref{fig:parameter-space-computations-instruments} shows the number of complex operations per frequency channel per cross-polarization per integration time-scale. Telescopes that fall to the left of the solid line indicate MOFF imaging is computationally more efficient than visibility-based imaging. All HERA layouts except HERA-19 and HERA-37 are in a region of parameter space where MOFF imaging holds the advantage. The dashed line showing future trajectory of HERA-like systems will be clearly favoured by MOFF imaging. The dotted line is similarly a hypothetical trajectory for the MWA with more tiles added inside the same core diameter. The grey shaded area is for a projected LWA expansion and is also predominantly in the region favouring MOFF imaging. It is bounded by the LWA1 and LWA-OV on the left and right respectively. The current (see Table~\ref{tab:antenna-layouts}) and a hypothetical expanded layout with a four-fold increase in number of elements over a 50\% increase in $b_\textrm{max}$ provide the bounds at the bottom and top respectively. Current instruments such as MWA and LOFAR lie in parameter space favouring visibility-based imaging {\bf due to the sparseness of their layouts}.

\begin{figure}
  \includegraphics[width=\columnwidth]{figure11}
  % \includegraphics[width=\columnwidth]{MOFF_FX_computations_fov_gridding_annotated}
  % \includegraphics[width=\columnwidth]{MOFF_FX_computations_fov_gridding_annotated.eps}
  \caption{Current and planned instruments in parameter space of number of complex multiplications and additions. The solid line is the boundary at which the number of operations with MOFF and visibility-based imaging are equal. MOFF imaging is more efficient for telescopes occupying the left of this line and vice versa. CHIME, HIRAX, SKA1-LC, SKA1-LCD and all the HERA layouts except HERA-19 and HERA-37 lie in the parameter space favoured by MOFF imaging. The dashed line shows the projected trajectory of hypothetical expanded HERA layouts. The dotted line similarly shows hypothetical expanded MWA layouts with more tiles added in the same core. The grey shaded area denotes the projected trajectory of the LWA bounded by LWA1 (left edge), LWA-OV (right edge), current layout (bottom) and a four-fold increase in the number of elements within a 50\% increase in the core size (top). Current instruments such as MWA and LOFAR fall in a region favoured by visibility-based imaging {\bf due to sparse layouts}.}
  \label{fig:parameter-space-computations-instruments}
\end{figure}

@@

As radio astronomy is entering a new era, advances in instrumentation have to be accompanied by equal advances in processing techniques to manage computational resources. Many future radio telescopes such as the SKA, HERA and LWA are headed towards the large-N dense array layout model for which computational cost from traditional FX/XF correlator-based architecture and visibility-based imaging starts rising steeply. We have provided the first software demonstration of a general purpose imaging algorithm using our generic and efficient EPIC software that is designed to bring this cost down from $\mathcal{O}(\Nant^2)$ to $\mathcal{O}(\Ngrid\log_2 \Ngrid)$. Under the class of direct imaging techniques, ours is one of the most generic -- neither does it place any constraint on the array layout to be on a regular grid nor does it require the antenna array to be homogeneous. {\bf We have provided a detailed mathematical framework for imaging with heterogeneous arrays, demonstrated the natural ability of EPIC} to robustly image data from such arrays, and shown that wrong assumptions about array homogeneity could result in significant and systematic mis-estimation of the sky model.

@@

For a heterogeneous array, it is necessary to keep track of antennas and antenna pairs of different types which will determine the final weighting in the synthesized image. We consider a heterogeneous array consisting of a total of $\Nant$ antennas under $N_\textrm{T}$ different types. And under each antenna type $p$, there are $n_p$ antennas such that $\sum_p n_p = \Nant$ {\bf (unrelated to the noise quantity also denoted by $n$ in \S\ref{sec:math})}. The total number of unique non-zero spacing antenna pairs is $\Nant(\Nant-1)/2$ and there are potentially up to $N_\textrm{T}(N_\textrm{T}+1)/2$ unique antenna pair types obtained by pairwise combination of the $N_\textrm{T}$ antenna types. The total number of antenna pairs, 
\begin{align}\label{eqn:apndx-total-baselines}
  \Nant(\Nant-1) &= \sum_{pq} n_{pq}, \\
  \textrm{where}, \quad   n_{pq} &= \begin{cases}
    n_p\,n_q, & p\ne q \\
    n_p(n_p-1), & p=q, 
  \end{cases}
\end{align}
is obtained by counting antenna pairs, $n_{pq}$, in different antenna pair types, $pq$. {\bf Note that $ab$ and $pq$ act as simple indices into pairs formed from individual indices and not their product.}

>>>>

% Only include extra packages if you really need them. Common packages are:
\usepackage{graphicx}	% Including figure files
\usepackage{amsmath}	% Advanced maths commands
\usepackage{amssymb}	% Extra maths symbols
\usepackage[flushleft]{threeparttable} % For table notes
\usepackage[caption=false]{subfig} % For subfloats

@@

% Abstract of the paper
\begin{abstract}
Modern radio telescopes are favouring densely packed array layouts with large numbers of antennas ($\Nant\gtrsim 1000$). Since the complexity of traditional correlators scales as $\mathcal{O}(\Nant^2)$, there will be a steep cost for realizing the full imaging potential of these powerful instruments. Through our generic and efficient E-field Parallel Imaging Correlator (EPIC), we present the first software demonstration of a generalized direct imaging algorithm, namely, the Modular Optimal Frequency Fourier (MOFF) imager. Not only does it bring down the cost for dense layouts to $\mathcal{O}(\Nant\log_2\Nant)$ but can also image from irregular layouts and heterogeneous arrays of antennas. EPIC is highly modular, parallelizable, implemented in object-oriented Python, and publicly available. We have verified the images produced to be equivalent to those from traditional techniques to within a precision set by gridding coarseness. We have also validated our implementation on data observed with the Long Wavelength Array (LWA1). We provide a detailed framework for imaging with heterogeneous arrays and show that EPIC robustly estimates the input sky model for such arrays. Antenna layouts with dense filling factors consisting of a large number of antennas such as LWA, the Square Kilometre Array, Hydrogen Epoch of Reionization Array, and Canadian Hydrogen Intensity Mapping Experiment will gain significant computational advantage by deploying an optimized version of EPIC. The algorithm is a strong candidate for instruments targeting transient searches of Fast Radio Bursts (FRB) as well as planetary and exoplanetary phenomena due to the availability of high-speed calibrated time-domain images and low output bandwidth relative to visibility-based systems.
\end{abstract}

@@

Here we present the first software implementation of the MOFF correlator, and announce the public release of the E-field Parallel Imaging Correlator (EPIC) code. EPIC is a highly parallel, object-oriented Python package that primarily implements the MOFF imaging algorithm, emulates real-life telescopes and FX/XF correlators in software, and includes a visibility-based imaging technique for reference. It is intended to provide a development platform to test different imaging approaches, characterize scaling relations and serve as a stepping stone for real-life GPU/FPGA-based implementation on telescopes.

We begin with a technical description of the algorithm in \S\ref{sec:math}, then discuss our particular implementation in \S\ref{sec:software}. We then verify the output data quality from our code in \S\ref{sec:verify} by presenting simulated images from both the EPIC correlator and comparing to a simulated FX correlator. We also demonstrate the performance with real-world data from the LWA1. In \S\ref{sec:versatility}, we present a mathematical framework for imaging with heterogeneous arrays and demonstrate the ability of EPIC to robustly image data from such arrays. In \S\ref{sec:analysis}, we explore the scalability of the algorithm in the context of several array design choices. We identify specific array design classes where the EPIC correlator is computationally more efficient, and in the field of transients, demands significantly lesser I/O bandwidth relative to visibility-based approaches. We conclude and discuss future research prospects in \S\ref{sec:conclusions}.

@@

We can then use the multiplication-convolution theorem to move the convolution in Equation~\ref{eqn:e-field-conv} to a product after the Fourier transform in equation~\ref{eqn:dirty-image-FX}.
\begin{align}
  \boldsymbol{\mathcal{I}}^\prime(\hat{\mathbf{s}}) &= \left\langle \left|\,\mathbf{F}^\textrm{T}(\hat{\mathbf{s}},\mathbf{r})\,\mathbf{W}^\textrm{T}(\mathbf{r},\mathbf{a})\,\widetilde{\mathbf{N}}(\mathbf{a},\mathbf{a})\,\mathbf{E}(\mathbf{a})\,\right|^2\right\rangle_t. \label{eqn:dirty-image-MOFF}
\end{align}
The term inside the angular brackets before squaring has a very similar form as that in equation~\ref{eqn:dirty-image-FX}. It signifies that the measured antenna electric fields are weighted by the antenna noise, weighted and gridded by the antenna aperture kernel, Fourier-transformed and finally squared to obtain the same image estimate that would have been obtained using equation~\ref{eqn:dirty-image-FX}.

@@

In a companion paper \citep{bea16}, we demonstrate a novel calibration technique (EPICal) which leverages the data products formed by direct imaging correlators to estimate antenna complex gains. This method correlates the antenna electric field signals with an image pixel from the output of the correlator in the feedback calibration fashion outlined in \citealt{mor11} (illustrated in Fig.~\ref{fig:MOFF-flowchart} by the arrow leading from the imager to the calibration block). Furthermore, it allows for arbitrarily complex sky models, places no restrictions on array layout, and accounts for non-identical antenna beam patterns. Direction-dependent calibration can be achieved by correlating antenna signals with output pixels in the direction of $N_\textrm{c}$ calibration sources, then fitting for a functional model of the sky. Since antennas are only correlated with calibrator pixels, the computational complexity scales as $\sim N_\textrm{a} N_\textrm{c}$. The calibration performance of EPIC is explored for varying levels of completeness of sky model in \citet{bea16}.

@@

In contrast, visibilities measured by an antenna are statistically stable within an integration time interval. Hence, they are averaged after (and sometimes before) calibration as shown in equation~\ref{eqn:cc-vis}. It is advantageous to average them in visibilities before imaging because the repeated cost of spatial FFT can be avoided. Since this averaging has been performed already on the visibilities over an integration time-scale, the imaging step has to be performed only once per integration cycle.

@@

We use the EPIC simulator to generate stochastic electric field samples from a sky model consisting of 10 point sources of random flux densities $>10$~Jy each at random locations. In our simulations, we use 64 frequency channels each of width $\Delta f = 40$~kHz. The number of time stamps integrated in one integration cycle was kept at eight where the duration of each time-series is $1/\Delta f=25\,\mu$s long. We use the MWA array layout \citep{bea12} for demonstration. Only the inner 51 tiles within a square bounding box of 150~m on each side were used. We assumed all tiles are identical and have a square-shaped electric field illumination footprint 4.4~m on each side. Besides the stochastic sky noise present in the simulated electric fields, no noise from the instrument is added.

@@

For a heterogeneous array, 
\begin{align}\label{eqn:wt-dirty-image-EPIC-decomp}
  \mathcal{I}^\prime &= \frac{1}{\Nant(\Nant-1)}\,\sum_{pq}\sum_{\substack{ab\\a\ne b\\a\in p, b\in q}}\,\mathcal{I}^\prime_{ab},
\end{align}
where, $ab$ indexes pairs of antennas $a$ and $b$, $pq$ indexes the different types of antenna pairs, with $p$ and $q$ indexing antenna types (note that $ab$ and $pq$ are simply indices in our notation and are not to be interpreted as products of the individual indices). Thus, equation~\ref{eqn:wt-dirty-image-homogeneous} is re-written as the sum of fringes over all unique antenna pairs in an antenna pair type (inner sum) and subsequently summed over all antenna pair types (outer sum).

For purposes of demonstration in this study, we assume that the spatial distribution of antenna pairs in each antenna pair type is similar and thus results in a similar dirty image, $\mathcal{I}_\textrm{D}^\textrm{iso}$. This usually holds when antennas under each antenna type are chosen randomly or identically such that differences in the synthesized beams are insignificant. Then, equation~\ref{eqn:wt-dirty-image-EPIC-decomp} can be approximated as:
\begin{align}\label{eqn:wt-dirty-image-EPIC-decomp-approx1}
  \mathcal{I}^\prime &\approx \frac{\mathcal{I}_\textrm{D}^\textrm{iso}}{\Nant(\Nant-1)}\,\sum_{pq}\sum_{\substack{ab\\a\ne b\\a\in p, b\in q}} \mathcal{W}^{\textrm{G}\star}_a\,\mathcal{W}^\textrm{G}_b\,\mathcal{W}^\textrm{I}_a\,\mathcal{W}^{\textrm{I}\star}_b,
\end{align}
where, $\mathcal{I}_\textrm{D}^\textrm{iso}$ is the same for all terms and has been pulled out of the summations. For all antennas indexed by $a$ that belong to a particular type $p$, we replace $\mathcal{W}_a$ with $\mathcal{W}_{(p)}$ and is applicable to those arising from both instrumental (superscript `$I$') and gridding (superscript `$G$') origins. If $n_{pq}$ is the number of antenna pairs in the antenna pair type $pq$ (unrelated to the noise quantity also denoted by $n$ in \S\ref{sec:math}), $\mathcal{W}_\textrm{eff}$ can be simplified as:
\begin{align}\label{eqn:wt-dirty-image-EPIC-decomp-approx2}
  \mathcal{W}_\textrm{eff} &\approx \frac{1}{\Nant(\Nant-1)}\,\sum_{pq}n_{pq}\,\mathcal{W}^{\textrm{G}\star}_{(p)}\,\mathcal{W}^\textrm{G}_{(q)}\,\mathcal{W}^\textrm{I}_{(p)}\,\mathcal{W}^{\textrm{I}\star}_{(q)}.
\end{align}
In an optimally weighted image, $\mathcal{W}^\textrm{G}_{(p)} \equiv \mathcal{W}^\textrm{I}_{(q)}$. Thus,
\begin{align}\label{eqn:effective-weighting-optimal}
  \mathcal{W}_\textrm{eff}^\textrm{opt} &\approx \frac{1}{\Nant(\Nant-1)}\,\sum_{pq} n_{pq}\,\left|\mathcal{W}^\textrm{I}_{(p)}\right|^2\left|\mathcal{W}^\textrm{I}_{(q)}\right|^2,
\end{align}
where, superscript `$\textrm{opt}$' denotes `optimal'.

@@

The number of complex multiplications and accumulations in the spatial Fourier transform implemented via FFT is $\approx \beta\,(4\Ngrid)\log_2(4\Ngrid)$ where $\Ngrid$ is the number of pixels on the grid, the factor 4 accounts for the increase in the number of pixels as a result of zero-padding before the spatial Fourier transform, and $\beta$ is a constant that depends on the implementation of twiddle FFT algorithms \citep{bri74}. In our study, we set $\beta=5$, a value\footnote{http://www.fftw.org/speed/method.html} much more conservative than was indicated in \citet{mor11}. We set the number of complex multiplications in the X-engine in visibility-based imaging to $\Nant(\Nant-1)/2$.

@@

Fig.~\ref{fig:parameter-space-computations-instruments} shows the number of complex operations per frequency channel per cross-polarization per integration time-scale. Telescopes that fall to the left of the solid line indicate MOFF imaging is computationally more efficient than visibility-based imaging. All HERA layouts except HERA-19 and HERA-37 are in a region of parameter space where MOFF imaging holds the advantage. The dashed line showing future trajectory of HERA-like systems will be clearly favoured by MOFF imaging. The dotted line is similarly a hypothetical trajectory for the MWA with more tiles added inside the same core diameter. The grey shaded area is for a projected LWA expansion and is also predominantly in the region favouring MOFF imaging. It is bounded by the LWA1 and LWA-OV on the left and right respectively. The current (see Table~\ref{tab:antenna-layouts}) and a hypothetical expanded layout with a four-fold increase in number of elements over a 50\% increase in $b_\textrm{max}$ provide the bounds at the bottom and top respectively. Current instruments such as MWA and LOFAR lie in parameter space favouring visibility-based imaging due to the sparseness of their layouts.

\begin{figure}
  \includegraphics[width=\columnwidth]{figure11}
  % \includegraphics[width=\columnwidth]{MOFF_FX_computations_fov_gridding_annotated}
  % \includegraphics[width=\columnwidth]{MOFF_FX_computations_fov_gridding_annotated.eps}
  \caption{Current and planned instruments in parameter space of number of complex multiplications and additions. The solid line is the boundary at which the number of operations with MOFF and visibility-based imaging are equal. MOFF imaging is more efficient for telescopes occupying the left of this line and vice versa. CHIME, HIRAX, SKA1-LC, SKA1-LCD and all the HERA layouts except HERA-19 and HERA-37 lie in the parameter space favoured by MOFF imaging. The dashed line shows the projected trajectory of hypothetical expanded HERA layouts. The dotted line similarly shows hypothetical expanded MWA layouts with more tiles added in the same core. The grey shaded area denotes the projected trajectory of the LWA bounded by LWA1 (left edge), LWA-OV (right edge), current layout (bottom) and a four-fold increase in the number of elements within a 50\% increase in the core size (top). Current instruments such as MWA and LOFAR fall in a region favoured by visibility-based imaging due to sparse layouts.}
  \label{fig:parameter-space-computations-instruments}
\end{figure}

@@

As radio astronomy is entering a new era, advances in instrumentation have to be accompanied by equal advances in processing techniques to manage computational resources. Many future radio telescopes such as the SKA, HERA and LWA are headed towards the large-N dense array layout model for which computational cost from traditional FX/XF correlator-based architecture and visibility-based imaging starts rising steeply. We have provided the first software demonstration of a general purpose imaging algorithm using our generic and efficient EPIC software that is designed to bring this cost down from $\mathcal{O}(\Nant^2)$ to $\mathcal{O}(\Ngrid\log_2 \Ngrid)$. Under the class of direct imaging techniques, ours is one of the most generic -- neither does it place any constraint on the array layout to be on a regular grid nor does it require the antenna array to be homogeneous. We have provided a detailed mathematical framework for imaging with heterogeneous arrays, demonstrated the natural ability of EPIC to robustly image data from such arrays, and shown that wrong assumptions about array homogeneity could result in significant and systematic mis-estimation of the sky model.

@@

For a heterogeneous array, it is necessary to keep track of antennas and antenna pairs of different types which will determine the final weighting in the synthesized image. We consider a heterogeneous array consisting of a total of $\Nant$ antennas under $N_\textrm{T}$ different types. And under each antenna type $p$, there are $n_p$ antennas such that $\sum_p n_p = \Nant$ (unrelated to the noise quantity also denoted by $n$ in \S\ref{sec:math}). The total number of unique non-zero spacing antenna pairs is $\Nant(\Nant-1)/2$ and there are potentially up to $N_\textrm{T}(N_\textrm{T}+1)/2$ unique antenna pair types obtained by pairwise combination of the $N_\textrm{T}$ antenna types. The total number of antenna pairs, 
\begin{align}\label{eqn:apndx-total-baselines}
  \Nant(\Nant-1) &= \sum_{pq} n_{pq}, \\
  \textrm{where}, \quad   n_{pq} &= \begin{cases}
    n_p\,n_q, & p\ne q \\
    n_p(n_p-1), & p=q, 
  \end{cases}
\end{align}
is obtained by counting antenna pairs, $n_{pq}$, in different antenna pair types, $pq$. Note that $ab$ and $pq$ act as simple indices into pairs formed from individual indices and not their product.
